Testing the Relationship between Phenomenological Control related to Illusion Sensitivity

### 2024-04-26 paper/manuscript.qmd
preprint

    # Example: This article is based on data published in [Reference].

@@

Interestingly, @lush2022 did test for a relationship between PC and illusion sensitivity using the Müller-Lyer illusion (in which the arrangement of the arrowheads flanking two lines makes them appear as having different lengths), and reported evidence in favour of an absence of correlation between the two measures. This finding was interpreted as indicative of the cognitive impenetrability of illusions, implying that the effect is driven by low-level processes and therefore not influenced by top-down mechanisms such as PC. The goal of this study is thus to replicate the results from @lush2022 pointing to an absence of a relationship between phenomenological control and illusion sensitivity, by generalising them to a different illusion task that would encompass other illusion types.

@@

We aim to recruit around 500 [in line with the sample sizes used in @lush2021; @lush2022] adult English native speakers with a desktop device using Prolific (www.prolific.co). Participants will be first presented with an explanatory statement and the consent form, and can proceed by pressing a button to confirm they have read and understood the information. This study will be approved by the ethics board of the School of Psychology of the University of Sussex (XX/XXXX).

@@

>>>>

@@

Interestingly, @lush2022 did test for a relationship between PC and illusion sensitivity using the Müller-Lyer illusion (in which the arrangement of the arrowheads flanking two lines makes them appear as having different lengths), and reported evidence in favour of an absence of correlation between the two measures. This finding was interpreted as indicative of the cognitive impenetrability of illusions, implying that the effect is driven by low-level processes and therefore not influenced by top-down mechanisms such as PC. The goal of this study is thus to replicate the results from @lush2022 pointing to an absence of a relationship between phenomenological control and illusion sensitivity, by generalising them to a different illusion task that would encompass other illusion types (see @tbl-table1).

::: {#tbl-table1}

| **Question**                                                                                                                                 | **Hypothesis**                                                                                                                            | **Sampling Plan**                                                                                                               | **Analysis Plan**                                                                                                                                                                             | **Rationale for Deciding the Sensitivity of the Test**                                                                                           | **Interpretation Given Different Outcomes**                         | **Theory That Could Be Shown Wrong by the Outcomes**                                                                                                        |
|----------------------------------------------------------------------------------------------------------------------------------------------|------------------------------------------------------------------------------------------------------------------------------------------|--------------------------------------------------------------------------------------------------------------------------------|------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|-------------------------------------------------------------------------------------------------------------------------------|--------------------------------------------------------------------|----------------------------------------------------------------------------------------------------------------------------------------------------|
| Is there a correlation between trait phenomenological control (PC) and visual illusion (VI) sensitivity?                                    | Replicating findings from Lush et al., 2022 paper, we expect evidence of favour of an absence of relationship between VI and PCS         | The goal is to recruit around 500 adult English speakers using Prolific. This sample size is based on the ones used in Lush et al., 2021 and Lush et al., 2022 that we aim at replicate. | Bayesian correlations between the PC score and the VI performance for the 3 illusion types (corresponding to the error rate) will be computed using the BayesFactor::BFCorrelation() function (with the r-scale prior parameter set to 'medium') | Evidence against a relationship between PC and VI will be found if BF10 <= 1/3, following the Lush et al., 2022 findings. BF10 > 3 will be interpreted as evidence for a relationship between these two measures. | The hypothesis that VI sensitivity is independent from PC.          | The cognitive impenetrability of illusions, implying that the effect is driven by low-level processes and therefore not influenced by top-down mechanisms such as PC. |

Study Design Table
:::

@@

We aim to recruit around 500 [in line with the sample sizes used in @lush2021; @lush2022] adult English native speakers with a desktop device using Prolific (www.prolific.co). Participants will be first presented with an explanatory statement and the consent form, and can proceed by pressing a button to confirm they have read and understood the information. This study has been approved by the ethics board of the School of Psychology of the University of Sussex (ER/ASF25/5).

@@

# Results

*This section will be completed after data is collected.*

# Discussion

*This section will be completed after data is collected.*

# Data Availability

All the study materials, experiment, data, and analysis is available on GitHub at https://github.com/RealityBending/IllusionGamePhenomenologicalControl

# Acknowledgments

We would like to thank An Shu Te for her help in setting up the project, Ryan Scott for his help in implementing the phenomenological control scale, and Zoltan Dienes for his input, feedback and guidance.
